Spike spectra for recurrences

### 2022-11-04 Manuscript/inter_spike_manuscript_mdpi.old.tex
revision

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{In recurrence analysis, the $\tau$-recurrence rate encodes the periods of the cycles of the underlying high-dimensional time series. It thus plays a similar role to the autocorrelation for scalar time-series in encoding temporal correlations. However, its Fourier decomposition does not have a clean interpretation. Thus there is no satisfactory analogue to the power spectrum in recurrence analysis. In this paper we introduce a novel way to decompose the $\tau$-recurrence rate using an over-complete basis of Dirac combs together with a sparsity regularization. We show that this decomposition, the \textit{inter-spike spectrum}, naturally provides an analogue to the power spectrum for recurrence analysis in the sense that it reveals the dominant periodicities of the underlying time series. We show that the inter-spike spectrum correctly identifies patterns and transitions in the underlying system in a wide variety of examples and is robust to measurement noise.}
%
%A novel kind of power spectrum is constructed, the \textit{inter-spike spectrum}, which transforms any signal into its spike-frequency domain. This method 
%clearly shows the apparent cycles in the data and overcomes the problems for spike-train-like signals when using the obvious idea of Fourier-transforming it. 
%We invent this instructive approach with the 
%idea of transforming the $\tau$-recurrence rate of a recurrence plot (RP), which often has a spiky appearance. The $\tau$-recurrence rate is the density of recurrence points 
%along diagonals of the RP, which are parallel to the main diagonal with a distance of $\tau$. In this context the inter-spike spectrum can be interpreted as a nonlinear power 
%spectrum of a potentially high dimensional system which constitutes the RP. The proposed measure is robust to noise and is able to detect and analyze bifurcations.}

@@

The dynamics of complex systems as provided by measured time series usually show complicated and chaotic patterns.
Quantifying their recurrence features is a powerful way to describe them and to infer information about 
the type of dynamics, stability, regime changes, or couplings and synchronisation \cite{marwan2007,marwan2008epjst,webber2015}.
Even more challenging are signals which do have a heavy tailed-distribution or appear as a spike-train,
e.g., neuron firings \cite{Dummer2014,Orcioni2020}, heart beat variability \cite{marwan2002herz}, 
or extreme flood events \cite{banerjee2021}.
Deriving useful information from spike-train signals or inter-spike time series is an important
topic in data analysis in many scientific fields \cite{Kajikawa2005,Dummer2014,Orcioni2020,Canale2021}.

Recurrence plots (RPs) provide a vivid representation of complex dynamics $\vec{x}_i$ stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N],
\end{equation}
with $\mathbf{R}$ the recurrence matrix, $\vec{x}_i$ the state vector at time 
$t = \Delta t \cdot i$ ($\Delta t$ the sampling time), $N$ the number of
sampling points (or length of data series), and $d$ the dimension of the system.
The crucial free parameter $\varepsilon$ is the recurrence threshold, determining what is a recurrences
and, thus, the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

Some of these recurrence quantification measures, the entropy of diagonal lines and the entropy of 
recurrence times, can be related to basic characteristics of complex systems, such as the Kolmogorov-Sinai entropy \cite{march2005,baptista2010}. However, these quantifiers have a free parameter, the minimal considered line length, and 
are usually biased, due to the finite size of the RP and thickened diagonal lines, which needs to be corrected \cite{Kraemer2019}. Moreover, the mentioned statistics cannot account for 
changing regular (non-chaotic) dynamics, such as period-doubling bifurcations.

A rather simple idea is to look at the $\tau$-recurrence rate of the RP ($\tau$-RR, Eq.~\ref{eq_tau_rr}) \cite{marwan2002pla,Zbilut2008}.
This is the density of recurrence points along the diagonals of the recurrence matrix, as a function of the distance $\tau$ (sampling units) to the main diagonal:
\begin{equation}\label{eq_tau_rr}
\tau\text{-RR}(\varepsilon) = RR(\tau, \varepsilon) = \frac{1}{N-\tau} \sum_{i=1}^{N-\tau	} R_{i,i+\tau}.
\end{equation}
$\tau$-RR serves as an estimator for the probability that the system recurs after time $\tau \Delta t$, with $\Delta t$ being the sampling time of the trajectory 
$\vec{x}_i = \vec{x}(\Delta t \cdot i),\, i=1,\ldots,N$. 
It represents the period length of cycles in the data (Fig.~\ref{fig_tau_rr_spectrum_example}D).

\citet{Zbilut2008} pointed out that $\tau$-RR could be interpreted as analogous to the auto-correlation function $C(\tau)$ and, hence, via the Wiener-Khinchin theorem, provide an analogue,
``generalized'', spectral density. This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR. There are clear advantages for a recurrence-derived spectral density, 
i.e., Fourier transforming (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E), instead 
of $C(\tau)$: There are no assumptions for stationarity or sampling, when constructing a RP.
Furthermore, since an RP can represent high-dimensional dynamics and its $\tau$-RR serves as a plug-in for $C(\tau)$, the correlation structures of higher dimensional spaces can be read from 
the recurrence-derived Fourier-spectrum.\\

@@

However, the interpretation of this generalized power spectral density is unclear, and it is typically hard to interpret. $\tau$-RR directly encodes the periodicity of the underlying signal, in contrast it is unclear what, if any, interpretation the "power" contained in a particular frequency mode of the $\tau$-RR should have. Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about the systems' state space trajectory.

For example, consider the signal we would like to analyze 
(e.g., the $\tau$-RR of a system) to be a Dirac comb (DC) with inter-spike period $T_\text{is}$: 
\begin{equation}
\text{DC}_{\text{is}}(t) = \sum_{k=-\infty}^{\infty} \delta(t-kT_\text{is}),
\label{eq_dirac_comb}
\end{equation}
i.e., a series of Dirac delta functions for a period $T_\text{is}$. There is only one single period, $T_\text{is}$, in this signal (Fig.~\ref{fig_tau_rr_dirac_comb}A, D), so 
in principle we would strive for a single peak in the frequency domain of this signal at a frequency $f=1/T_\text{is}$. The Fourier spectrum does not meet this expectation 
and instead of a single frequency, there are many frequencies excited (Fig.~\ref{fig_tau_rr_dirac_comb}B, E). This is because the Fourier components add constructively 
for every frequency $1/T_\text{is}$ and therefore $\text{DC}_{\text{is}}(t)$ coincides with its own Fourier transform up to a factor $1/T_\text{is}$ \cite{Norden1998}.\footnote{This can also 
be observed for neuron spike trains, e.g. \cite{Orcioni2020,Biagetti2017}.}

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_dirac_comb}
 \caption{The transformation of a Dirac comb (series of Dirac delta functions) with a single inter-spike period $T_{\text{is}}=100$ ($\widehat{=}f=0.01$) 
 into the frequency domain. 
 \textbf{A} Dirac Comb (DC) with equal amplitudes and
 \textbf{B} its FFT-based power spectral density.
 \textbf{C} Proposed inter-spike spectrum of the signal in \textbf{A} showing a single frequency, which corresponds to the inter-spike period $T_{\text{is}}$ ($f=0.01$).
 \textbf{D} DC with randomly chosen amplitudes and same $T_{\text{is}}$ as in \textbf{A}, and
 \textbf{E} its FFT-based power spectral density. 
 \textbf{F} Proposed inter-spike spectrum of the signal in \textbf{D} showing a single frequency, which corresponds to the expected inter-spike period $T_{\text{is}}$ ($f=0.01$). 
 Inter-spike spectra were obtained with a LASSO regression and a regularization threshold corresponding to $\rho=0.9$ accordance of the signals in (\textbf{A},\textbf{D}) and 
 its re-composed signals (c.f. Section~\ref{sec_tau_rr_method}). 
}
\label{fig_tau_rr_dirac_comb}
\end{figure}

In this article we propose a new way of decomposing a spike-train-like signal into periodic components as 
an alternative to the RP-based method suggested in \cite{Zbilut2008} or Fourier-based spike-train power 
spectra \cite{Dummer2014}. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies in such a way that harmonics of these frequencies are also excited (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea. Note that this approach can be used to decompose arbitrary signals, and is not specific to $\tau$-RR. However, the more spiky the signal is, the more useful our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we demonstrate the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

The signal, which we would like to transform\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} is decomposed into a set of appropriate basis functions. 
The general idea is that the sum of these weighted functions can approximate a finite signal to a sufficient degree. The weights (also called modes) corresponding to the individual basis functions can also 
be called \textit{loadings} and must be determined. A number of decomposition techniques are out there based on different sets of basis functions, e.g., 
trigonometric functions (Fourier- and wavelet analysis \cite{Bracewell1986}), 
eigenvectors of the corresponding covariance matrix (Principal Component Analysis \cite{Hotelling1933} and related techniques) 
or Intrinsic Mode Functions (Empirical Mode Decomposition and Hilbert spectrum \cite{Norden1998}).

Here we propose to use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the min-max-normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad \text{such that}\quad n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

The proposed decomposition does not show a leakage effect by construction. However, when introducing a sparsity regularization we give up the ability to fully reconstruct the signal from the decomposition in favor of identifying the most important periodicities more clearly. This choice is mediated by a regression regularization parameter $\alpha$. Only for $\alpha = 0$ do we obtain a solution to Eq.~(\ref{eq_linear_system}) and thus a full reconstruction of the signal. Our general approach to fixing $\alpha$ is to require that we preserve most of the original signal in the sense of matching a given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and the reconstruction. The chosen regression method (in our case, LASSO or STLS) will determine how sensitive the obtained inter-spike spectrum is with respect to the regularization given by a 
selected $\rho$ (Fig.~\ref{fig_regularization}). Regarding this, in most of our applications LASSO yields more robust results, while STLS is usually faster to compute. 
We will discuss this in more detail in Section 4. Section \ref{sec_tau_rr_discussion}.

%
%In addition to the computational complexity of solving Eq.~\eqref{eq_linear_system}, the solving algorithm must also deal with the free choice of the regression regularization parameter $\alpha$, 
%which determines the sparsity of the regression coefficient vector $\hat{\boldsymbol\beta}$. Obviously, there will be numerous peaks in the inter-spike spectrum of a given signal, when $\alpha$ is small, 
%since $\alpha=0$ corresponds to the least squares solution and the corresponding $\hat{\boldsymbol\beta}$ is an exact solution of Eq.~(\ref{eq_linear_system}) and not sparse. On the other hand, 
%only a sparse number of non-zero loadings in $\hat{\boldsymbol\beta}$, i.e., peaks in the inter-spike spectrum, are obtained when $\alpha$ is sufficiently large. 
%The actual implementation of the regularization depends on the regression method chosen (in our case, LASSO or STLS), and consequently a given value of $\alpha$ will generally result in 
%different sparse loading vectors $\hat{\boldsymbol\beta}$. Therefore, in this paper we select $\alpha$ such that the re-composed signal $\tilde{\vec{s}}=\bfX^T\hat{\boldsymbol\beta}$ matches a 
%given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and itself.

@@

We demonstrate the use of the inter-spike spectrum in combination with the $\tau$-RR as outlined in 
Section \ref{sec_tau_rr_intro} on several interesting research questions. The procedure is the following:
\begin{itemize}[noitemsep]
\item[(1)] Compute a RP of the trajectory of the system, Eq.~\eqref{eq_rp_definition}. If only univariate data is available, perform a state space reconstruction for obtaining the trajectory first.
\item[(2)] Compute the $\tau$-RR of that RP, Eq.~\eqref{eq_tau_rr}.
\item[(3)] Transform the $\tau$-RR into the proposed inter-spike spectrum, see Section \ref{sec_tau_rr_method}.
\end{itemize}

@@

The inter-spike spectra unravel the specific dynamics, which are also apparent in the state space portraits (Fig.~\ref{fig_tau_rr_example_roessler}A, B, C) and in the 
$\tau$-RRs (Fig.~\ref{fig_tau_rr_example_roessler}G, H, I). The proposed idea is also 
robust to noise (see Fig.~\ref{fig_tau_rr_example_roessler_noise} in the Appendix). This is because the peaks of the $\tau$-RR are insensitive to noise. 
Figure \ref{fig_roessler_peaks} illustrates that 
while the peak shape does change in the presence of noise, its position does not, and this is what the inter-spike spectrum encrypts after all 
(see Fig.~\ref{fig_roessler_spectra_distances} in the Appendix for further analysis).

\begin{figure}
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_roessler_different_noise_levels_peaks}
 \caption{Peak positions of the obtained inter-spike spectra of the $\tau$-RRs for additive noise levels up to 50\% for the discussed R\"ossler dynamics, 
 \textbf{A} period-2 limit-cycle, \textbf{B} period-3 limit-cycle and \textbf{C} chaos. The size of the plotted markers scale with the detected peak height. The noise-free spectra are shown in Fig.~\ref{fig_tau_rr_example_roessler}J, K, L and 
 an example of these spectra with 5\% additive noise is shown in Fig.~\ref{fig_tau_rr_example_roessler_noise}J, K, L. Spectra were obtained with a LASSO regression and a regularization threshold 
  corresponding to $\rho=0.95$ accordance of $\tau$-RRs and re-composed signals.}
\label{fig_roessler_peaks}
\end{figure}

@@

\acknowledgments{
All computations have been carried out in MATLAB\textsuperscript{\textregistered} and the Julia language 
and made use of the packages \textit{DynamicalSystems.jl} \cite{datseris2018}, \textit{DifferentialEquations.jl} \cite{rackauckas2017}, \textit{Distances.jl} and 
\textit{OptimalTransport.jl} \cite{Zhang2022}.
}

>>>>

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerospace, agriculture, agriengineering, agronomy, ai, algorithms, allergies, alloys, analytica, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, currophthalmol, cyber, dairy, data, dentistry, dermato, dermatopathology, designs, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, ent, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, futureinternet, futureparasites, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, hazardousmatters, healthcare, hearts, hemato, heritage, highthroughput, histories, horticulturae, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijtm, ijtpp, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jzbg, kidney, kidneydial, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology, pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, sports, standards, stats, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{A novel kind of power spectrum is constructed, the \textit{inter-spike spectrum}, which transforms any signal into its spike-frequency domain. This method 
clearly shows the apparent cycles in the data and overcomes the problems for spike-train-like signals when using the obvious idea of Fourier-transforming it. 
We invent this instructive approach with the 
idea of transforming the $\tau$-recurrence rate of a recurrence plot (RP), which often has a spiky appearance. The $\tau$-recurrence rate is the density of recurrence points 
along diagonals of the RP, which are parallel to the main diagonal with a distance of $\tau$. In this context the inter-spike spectrum can be interpreted as a nonlinear power 
spectrum of a potentially high dimensional system which constitutes the RP. The proposed measure is robust to noise and is able to detect and analyze bifurcations.}

@@

Recurrence Plots (RPs) provide a vivid representation of complex dynamics stemming from potentially high dimensional systems \cite{marwan2007}
\begin{equation}\label{eq_rp_definition}
R_{i,j}(\varepsilon) = \Theta\left(\varepsilon - D_{i,j}\right) 
= \Theta\left(\varepsilon - \| \vec{x}_i - \vec{x}_j\|\right), \qquad \vec{x} \in \mathbb{R}^d, \quad i,j \in [1,\ldots, N].
\end{equation}
The crucial free parameter $\varepsilon$ (recurrence threshold) determines the visible structures in the RP. It can be chosen such that the recurrence rate 
$RR(\varepsilon)=N^{-2}\sum_{i,j}^N R_{i,j}(\varepsilon)$ exhibits a certain value \cite{kraemer2018}.
The simple idea to track recurring states of the $d$-dimensional trajectory $\vec{x}_i$ of the system under study not only allows for a beneficial visualization of the dynamics, 
but also for its 
quantification, using certain structures in the RP, such as diagonal or vertical lines \cite{marwan2007}.

Some of these recurrence quantification measures, the entropy of diagonal lines and the entropy of 
recurrence times, can be related to basic characteristics of complex systems, such as the Kolmogorov-Sinai entropy \cite{march2005,baptista2010}. However, these quantifiers have a free parameter, the minimal considered line length, and 
are usually biased, due to the finite size of the RP and thickened diagonal lines, which needs to be corrected \cite{Kraemer2019}. Moreover, the mentioned statistics cannot account for 
changing regular (non-chaotic) dynamics, such as period-doubling bifurcations.
A rather simple idea is to look at the $\tau$-recurrence rate of the RP ($\tau$-RR, Eq.~\ref{eq_tau_rr}) \cite{marwan2002pla,Zbilut2008}.
This is the density of recurrence points along the diagonals of the recurrence matrix, as a function of the distance $\tau$ (sampling units) to the main diagonal:
\begin{equation}\label{eq_tau_rr}
\tau\text{-RR}(\varepsilon) = RR(\tau, \varepsilon) = \frac{1}{N-\tau} \sum_{i=1}^{N-\tau	} R_{i,i+\tau}.
\end{equation}
$\tau$-RR serves as an estimator for the probability that the system recurs after time $\tau \Delta t$, with $\Delta t$ being the sampling time of the trajectory 
$\vec{x}_i = \vec{x}(\Delta t \cdot i),\, i=1,\ldots,N$. 
\citet{Zbilut2008} pointed out that $\tau$-RR could be used as a plugin value for the auto-correlation function $C(\tau)$ and, hence, via the Wiener-Khinchim theorem a 
``generalized'' powerspectrum can be obtained. This is reasonable, since the average distances for a given lag $\tau$ 
\begin{equation}
\overline{D}(\tau) = \frac{1}{N-\tau}\sum_{i=1}^{N-\tau} D_{i, i+\tau}
\end{equation}
can be directly read from the distance matrix $\mathbf{D}$ and is also preserved in its thresholded version $\tau$-RR. There are clear advantages for a recurrence-derived 
powerspectrum, i.e., Fourier transforming (FT) $\tau$-RR (Fig.~\ref{fig_tau_rr_spectrum_example}D,E), instead 
of $C(\tau)$: There are no assumptions for stationarity or sampling, when constructing a RP.
Furthermore, the correlation structures of higher dimensional spaces can be resolved in the recurrence-derived Fourier-spectrum.\\

@@

However, there is a drawback to this approach. Whenever $\tau$-RR is a spike-train-like signal, which it is in most cases (see Fig.~\ref{fig_tau_rr_spectrum_example}) especially for 
map-data (low-resolution data), a FT of such a signal leads to a spike-train-like image in the frequency domain (e.g., \cite{Schild1982,Cordoba1989}, see 
Fig.~\ref{fig_tau_rr_spectrum_example}E). Thus, it is not intuitive how to extract meaningful information about dominant frequencies of the systems' state space trajectory.

For clarification, consider the signal we would like to analyze 
(e.g., the $\tau$-RR of a system) to be a Dirac comb (DC) with inter-spike period $T_\text{is}$: 
\begin{equation}
\text{DC}_{\text{is}}(t) = \sum_{k=-\infty}^{\infty} \delta(t-kT_\text{is}),
\label{eq_dirac_comb}
\end{equation}
i.e., a series of Dirac delta functions for a period $T_\text{is}$. There is only one single period, $T_\text{is}$, in this signal (Fig.~\ref{fig_tau_rr_dirac_comb}A, D), so 
in principle we would strive for a single peak in the frequency domain of this signal at a frequency $f=1/T_\text{is}$. Surprisingly the Fourier spectrum does not meet this expectation 
and instead of a single frequency, there are exceptionally many frequencies excited (Fig.~\ref{fig_tau_rr_dirac_comb}B, E). This is because the Fourier components add constructively 
for every frequency $1/T_\text{is}$ and therefore $\text{DC}_{\text{is}}(t)$ coincides with its own Fourier transform up to a factor $1/T_\text{is}$.

\begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth]{./figures/fig_tau_rr_dirac_comb}
 \caption{The transformation of a Dirac comb (series of Dirac delta functions) with a single inter-spike period $T_{\text{is}}=100$ ($\widehat{=}f=0.01$) 
 into the frequency domain. 
 \textbf{A} Dirac Comb (DC) with equal amplitudes and
 \textbf{B} its FFT-based powerspectrum.
 \textbf{C} Proposed inter-spike spectrum of the signal in \textbf{A} showing a single frequency, which corresponds to the inter-spike period $T_{\text{is}}$ ($f=0.01$).
 \textbf{D} DC with randomly chosen amplitudes and same $T_{\text{is}}$ as in \textbf{A}, and
 \textbf{E} its FFT-based powerspectrum. 
 \textbf{F} Proposed inter-spike spectrum of the signal in \textbf{D} showing a single frequency, which corresponds to the expected inter-spike period $T_{\text{is}}$ ($f=0.01$). 
 Inter-spike spectra were obtained with a LASSO regression and a regularization threshold corresponding to $\rho=0.9$ accordance of the signals in (\textbf{A},\textbf{D}) and 
 its re-composed signals (c.f. Section~\ref{sec_tau_rr_method}). 
}
\label{fig_tau_rr_dirac_comb}
\end{figure}

In this article we propose a new way of transforming a spike-train-like signal into its frequency domain. This novel \textit{inter-spike spectrum} does not show resonance 
behavior of the signal's inherent inter-spike frequencies (Fig.~\ref{fig_tau_rr_dirac_comb}C, F). Section \ref{sec_tau_rr_method} explains the idea, which can be used to decompose 
any arbitrary signal and is not restricted to the $\tau$-RR, which we exemplify here. However, the more spiky the signal is, the more outstanding our new approach is compared to a 
FT. In Section \ref{sec_tau_rr_application} we exemplify the usage of inter-spike spectrum when transforming the $\tau$-RR of a system under study. In this case, the inter-spike spectrum 
can unravel characteristic time scales of high dimensional systems, which is not possible when using a FT. Finally, in Section \ref{sec_tau_rr_conclusion} our results are summarized.

@@

The signal, which we would like to transform is decomposed into a set of appropriate basis functions.\footnote{We focus on $\tau$-RR in this article, Eq.~\eqref{eq_tau_rr}, since 
such time series are often spiky, which favors the proposed decomposition technique. But the proposed method can be applied to any sort of signal.} Instead of using trigonometric 
functions, as it is the idea in the Fourier decomposition, we use Dirac combs (DC) with 
different inter-spike periods as basis functions, Eq.~\eqref{eq_dirac_comb}. Let $s(t_i)$ be the normalized signal we want to transform of length $N$ and 
$t_i=i\cdot \Delta t,~i=1,\ldots,N$, where $\Delta t$ denotes the sampling time and $s(t_i) \in [0,\ 1]\ \forall\ i$. In the following we label this time series as a $(1\times N)$-dimensional 
vector $\vec{s}$. 
First, $\tilde{N}$ different DCs of length $N$ are constructed with inter-spike periods $T_\text{is} \in [1,\ldots,\tilde{N}]$ and $\tilde{N}=\ceil*{N/2}+1$. Second, in order to account 
for possible phase shifts of 
these basis functions occurring in $\vec{s}$, each of these $\tilde{N}$ different DCs also need to be shifted one step further $T_\text{is}-1$ times. This leaves us with a total number of 
$M = \sum_{i=1}^{\tilde{N}}i$ 
basis functions which we can arrange as rows of a $(M\times N)$-sized matrix $\bfX$
(Fig.~\ref{fig_tau_rr_basis_functions} illustrates the described procedure):
\begin{align}
\label{eq_basis_matrix} 
\bfX_{i,j} &= \sum_{k=0}^{N}\delta\bigl(j-1-kT_\text{is}(i)-i+T_\text{is}(i) \bigr), \quad i=1,\dots,\ceil*{N/2}+1, \quad j=1,\dots,N\\
\label{eq_T_is} T_\text{is}(i)&= n, \quad s.t.~~ n:~ \frac{n(n-1)}{2} + 1 \leq i <  \frac{n(n+1)}{2} + 1,\quad n \in \mathbb{N}_+.
\end{align}

@@

In addition to the computational complexity of solving Eq.~\eqref{eq_linear_system}, the solving algorithm must also deal with the free choice of the regression regularization parameter $\alpha$, 
which determines the sparsity of the regression coefficient vector $\hat{\boldsymbol\beta}$. Obviously, there will be numerous peaks in the inter-spike spectrum of a given signal, when $\alpha$ is small, 
since $\alpha=0$ corresponds to the least squares solution and the corresponding $\hat{\boldsymbol\beta}$ is an exact solution of Eq.~(\ref{eq_linear_system}) and not sparse. On the other hand, 
only a sparse number of non-zero loadings in $\hat{\boldsymbol\beta}$, i.e., peaks in the inter-spike spectrum, are obtained when $\alpha$ is sufficiently large. 
The actual implementation of the regularization depends on the regression method chosen (in our case, LASSO or STLS), and consequently a given value of $\alpha$ will generally result in 
different sparse loading vectors $\hat{\boldsymbol\beta}$. Therefore, in this paper we select $\alpha$ such that the re-composed signal $\tilde{\vec{s}}=\bfX^T\hat{\boldsymbol\beta}$ matches a 
given (Pearson) correlation coefficient $\rho$ between the original signal $\vec{s}$ and itself.

@@

We exemplify the use of the inter-spike spectrum in combination with the $\tau$-RR as outlined in 
Section \ref{sec_tau_rr_intro} on several interesting research questions. The procedure is the following:
\begin{itemize}[noitemsep]
\item[(1)] Compute a RP of the trajectory of the system, Eq.~\eqref{eq_rp_definition}. If only univariate data is available, perform a state space reconstruction for obtaining the trajectory first.
\item[(2)] Compute the $\tau$-RR of that RP, Eq.~\eqref{eq_tau_rr}.
\item[(3)] Transform the $\tau$-RR into the proposed inter-spike spectrum, see Section \ref{sec_tau_rr_method}.
\end{itemize}

@@

The inter-spike spectra unravel the specific dynamics, which are also apparent in the state space portraits (Fig.~\ref{fig_tau_rr_example_roessler}A, B, C) and in the 
$\tau$-RRs (Fig.~\ref{fig_tau_rr_example_roessler}G, H, I). The proposed idea is also 
robust to noise (see Fig.~\ref{fig_tau_rr_example_roessler_noise} in the Appendix). This is because the peaks of the $\tau$-RR are insensitive to noise. While the peak shape does 
change in the presence of noise, its position does not, and this is what the inter-spike spectrum encrypts after all.

@@

\acknowledgments{
All computations have been carried out in MATLAB\textsuperscript{\textregistered} and the Julia language 
and made use of the packages \textit{DynamicalSystems.jl} \cite{datseris2018} and \textit{DifferentialEquations.jl} \cite{rackauckas2017}.
}
